---
title: "Genomic_offset_GF"
author: "Thomas Francisco"
date: "2024-05-15"
output: html_document
---
```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	cache = FALSE
)
library(ggplot2)
library(gradientForest)
library(pdist)
library(corrplot)
library(dplyr)
library(rnaturalearth)
```

```{r meta data}
meta_data_pop <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Populations/taxus_sample_29pop.csv",h=T,sep=";",dec=",")

#alphabetic order
meta_data_pop_order <- meta_data_pop[order(meta_data_pop$Population),]
```

This script aims to calculate the genomic offset using gradient forest. We already explained in the script RDA_genomic_offset the principles of genomic offset. As a summary, to calculate genomic offset, we need:  
    - past/present and future climatic data  
    - genomic data (optional: outliers)
    - calculate GEA relationship between genomic data (or outliers) and past/present 
    - interpolate/extrapolate the GEA relationship to space and time and calculate GO
    - Compare GO results across methods
    
    
Here, the GEA relationship will be estimate using gradient forest machine learning method (see explanations at the GF_candidate_selection script).

      1. Climatic data
      
Climatic data needs to be in a dataframe:
```{r load Past climatic data}
#Past climatic data
past_climatic <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Climatic_data/new_selection/Past_new_6_Climatic_data_scale_df.csv",sep=";",dec=",")


vars <- colnames(past_climatic[,-c(1:2)])
```

Present climatic data: 
```{r load present climatic data}
#Present climatic data
present_climatic <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/validation_GO/climatic_data/Present_climatic_data_T_adapcon_gentree_scaled.csv",sep=";",dec=",")
```


Future climatic data: 
 
```{r future clim data}
future_climatic <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Climatic_data/new_selection/Future_climatic_data_T_adapcon_gentree_scaled.csv",sep=";",dec=",")
```
 
      2. Genomic data
      
```{r genomic data}
#genomic data
load("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/genetic_new/data_allelic_frequencies_29pop_adapcon_gentree_475_8616.Rdata")

genomic_matrix <- data_allelic_frequencies_29pop_adapcon_gentree_475_8616
```
We can also load the sets of outliers retained:

- One with random SNPs
```{r load random SNPs set}
load("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/GEA_new_var/outliers/random_neutral_set_SNPs_T_adapcon_gentree.Rdata")

list_random_SNPs <- colnames(random_neutral_set_SNPs_T_adapcon_gentree)
```

- One with less conservatives thresholds    
```{r load outliers LC}
#set of less conservative thresholds
load("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/GEA_new_var/outliers/outliers_set_final_overlapping_no_LD_LC_new_var.Rdata")
```

  - One with more conservatives thresholds  
```{r load outliers MC}
load("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/GEA_new_var/outliers/outliers_set_final_overlapping_no_LD_new_var.Rdata")
```

      
      3. GEA relationship

The next step is to calculate the relationship between the outliers and climatic variables using the non linear model of GF. 
It's the same principle as the candidate selection with GF but here, we will summarize the info at the  turnover function for each snps using all predictors to have one turnover function for all the response var.

```{r GEA using gradient forest random}
#Run_GEA_GF_random <- gradientForest(data.frame(past_climatic[,vars],genomic_matrix[list_random_SNPs]), 
#                             predictor.vars=vars, 
#                             response.vars=colnames(genomic_matrix[list_random_SNPs]),
#                             corr.threshold=0.5, ntree=500, trace=T)
#
#length(Run_GEA_GF_random$result)
```


```{r save GEA GF}
#save(Run_GEA_GF_random,file="C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/GF/Run_GEA_GF_random.Rdata")
```

```{r load GEA GF}
load("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/GF/Run_GEA_GF_random.Rdata")
```
We save the output of GF because it can slightly change between runs



```{r GEA using gradient forest LC}
#Run_GEA_GF_LC <- gradientForest(data.frame(past_climatic[,vars],genomic_matrix[outliers_set_final_overlapping_no_LD_LC_new_var]), 
#                             predictor.vars=vars, 
#                             response.vars=colnames(genomic_matrix[outliers_set_final_overlapping_no_LD_LC_new_var]),
#                             corr.threshold=0.5, ntree=500, trace=T)

#length(Run_GEA_GF_LC$result)


```

We can see that 83 out of the 98 outliers are associated to some of the predictors. These 83 SNPs, will be used by GF in the next step to estimate Genomic offset.  


```{r save GEA GF}
#save(Run_GEA_GF_LC,file="C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/GF/Run_GEA_GF_LC.Rdata")
```

```{r load GEA GF}
load("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/GF/Run_GEA_GF_LC.Rdata")
```

```{r GEA using gradient forest MC}

#Run_GEA_GF_MC <- gradientForest(data.frame(past_climatic[,vars],genomic_matrix[outliers_set_final_overlapping_no_LD_new_var]), 
#                             predictor.vars=vars, response.vars=colnames(genomic_matrix[outliers_set_final_overlapping_no_LD_new_var]),
#                             corr.threshold=0.5, ntree=500, trace=T)

#length(Run_GEA_GF_MC$result)

#save(Run_GEA_GF_MC,file="C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/GF/Run_GEA_GF_MC.Rdata")
load("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/GF/Run_GEA_GF_MC.Rdata")
```



      4. Interpolate/extrapolate the relationship and calculate genomic offset
      
Now that we have our GEA model, we can interpolate/extrapolate the relationship to space/time to estimate the past and future genomic composition to calculate the genomic offset metric. To do that, we modified the github function of Fitzpatrick (url: https://github.com/fitzLab-AL/geneticOffsetR/blob/main/poplarGBS.gf.supportFunctions.R). 
      
```{r genomic offset function}
########## calculate adaptive offset for populations in space or time
genomic_offset_function <- function(gfMod, vars, env2, combined=F,
                       pops = envPop$pop_code, weighted=FALSE){
  
  #gfMod = gf model for prediction
  #vars = names of env variables
  #env2 = new environment (new place / time)
    transEnv2 <- predict(gfMod, env2[,vars]) #new env
    transEnv1 <- predict(gfMod) #current env
  
  #calculate Euclidean distance in transformed env space
  num <- nrow(transEnv1)
  dOut <- lapply(1:num, function(x, tEnv1, tEnv2){
    as.numeric(pdist(tEnv1[x,],  tEnv2[x,])@dist)}, tEnv2=transEnv2, tEnv1=transEnv1)
  return(dOut)
}
```

We can apply this function to our dataset: 

```{r Run genomic offset function}
list_set <- c("Run_GEA_GF_random","Run_GEA_GF_LC","Run_GEA_GF_MC")
list_period <- c("present", "future")

#vars = past climatic data used in GEA

for(i in 1: length(list_period)){
  
  period <- list_period[i]


for(x in 1:length(list_set)){
  GF_run <- get(list_set[x])
  name <- c("random","LC","MC")
  
  
  climate_to_calculate_GO <- get(paste0(period,"_climatic"))
  
Genomic_offset <- genomic_offset_function(gfMod=GF_run, vars=vars, env2=climate_to_calculate_GO[,vars], combined=F,
                       pops = row.names(genomic_matrix), weighted=FALSE)

#extraction GO values
Genomic_offset$values <- unlist(Genomic_offset)

genomic_offset_GF <- data.frame(Population=row.names(genomic_matrix),GO=Genomic_offset$values)
names(genomic_offset_GF)[2] <- paste0("genomic_offset_GF_",name[x])

assign(paste0("genomic_offset_GF_",name[x],"_",period),genomic_offset_GF)
  }
}
```

Based on these genomic offset values, we can plot such as for RDA GO, the values of each population: 

```{r plot GO values}
list_period <- c("present","future")
list_set <- c("random","LC","MC")

for(i in 1:length(list_period)){
  
  period <- list_period[i]
  
  for(x in 1: length(list_set)){
    
    set <- list_set[x]
GO_df <- get(paste0("genomic_offset_GF_",name[x],"_",period))
    
     #first, we need to add the coordinates
Genomic_offset_coord <- merge(GO_df,meta_data_pop_order[,c(2,4,5)],"Population")

#transform longitude and latitude to numeric variables
Genomic_offset_coord <- Genomic_offset_coord %>% mutate(Longitude=as.numeric(Longitude),Latitude=as.numeric(Latitude))


colors <- c( "darkgreen", "#FDF7B3","#FC4E2A","#BD0026","darkorchid4")
#background map
admin <- ne_countries(scale = "medium", returnclass = "sf")
      
plot <- ggplot(data = Genomic_offset_coord) + 
  geom_sf(data = admin, fill = gray(0.92), size = 0) +
  geom_point(aes(x = Longitude, y = Latitude, fill = cut_number(Genomic_offset_coord[,2], n = 5)), shape = 21,size=3, color = "black") +
  scale_fill_manual(
    values = colors,
    labels = c("low values","","","","high values"),
    drop = FALSE,na.translate = FALSE)+  # Ensure all levels are shown in the legend
  geom_sf(data = admin, fill = NA, size = 0.1) +
  coord_sf(xlim = c(-10, 30), ylim = c(35, 62), expand = FALSE) +
  xlab("Longitude") + ylab("Latitude") +
  guides(fill = guide_legend(title = "Genomic offset")) +
  ggtitle(paste0("Genomic offset across populations ",set," ",period)) +
  theme_bw(base_size = 11) +
  theme(legend.position="right", panel.grid = element_blank(), strip.text = element_text(size=11),plot.title = element_text(hjust = 0.5,color = "Black",face="italic"))
      
print(plot) 

#save
     pdf(paste0("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/GF/figures/GO/Genomic_offset_across_populations_",set,"_",period,".pdf"));print(plot);dev.off()
    
  }
}
```

We can see that there is not a clear pattern of genomic offset, it's not the more continental populations or along the coast or whatever patterns that exhibit higher genomic offset. We thought that maybe it could be a pattern of elevation with population in higher altitude exhibit higher genomic offset. 

```{r correlation GO elevation}
#load elevation

elevation_data <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Climatic_data/ClimateDT_extraction/extraction_climateDT_29pop_T.csv",h=T,sep=";",dec=",")

#merge elevation data with GO
data_GO_altitude <- data.frame(merge(genomic_offset_GF_LC_present,elevation_data,"Population"))
data_GO_altitude_df <- data_GO_altitude %>% 
  mutate(Longitude=as.numeric(Longitude),Latitude=as.numeric(Latitude),elevation=as.numeric(Elevation.DEM_90m.))

#df
df_tot_GO_predictions <- data.frame(Population=genomic_offset_GF_LC_future[,1],GO_rand_pres=genomic_offset_GF_random_present[,2],GO_rand_Fut=genomic_offset_GF_random_future[,2],GO_LC_pres=genomic_offset_GF_LC_present[,2],GO_LC_Fut=genomic_offset_GF_LC_future[,2],GO_MC_pres=genomic_offset_GF_MC_present[,2],GO_MC_Fut=genomic_offset_GF_MC_future[,2],elevation=data_GO_altitude_df$elevation)

#elevation

correlation <- cor(df_tot_GO_predictions[,-1])

corrplot(correlation, method = "number", addrect = 2, col = c("darkorange","darkred"), type = "lower", tl.col = "black", tl.cex = 0.6, number.cex = 0.6)


```

```{r save corrplot}
pdf("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/GF/figures/comparison/correlation_GO_values_GF_T_set_period.pdf");corrplot(correlation, method = "number", addrect = 2, col = c("darkorange","darkred"), type = "lower", tl.col = "black", tl.cex = 0.6, number.cex = 0.6);dev.off()
```
We can see that there is a small but present positive correlation between genomic offset and elevation. To see that, we will plot the elevation to the points in the genomic offset map: 

```{r plot GO and altitude}

colors <- c( "darkgreen", "#FDF7B3","#FC4E2A","#BD0026","darkorchid4")
#background map
admin <- ne_countries(scale = "medium", returnclass = "sf")
plot <- ggplot(data = data_GO_altitude_df) + 
  geom_sf(data = admin, fill = gray(0.92), size = 0) +
  geom_point(aes(x = Longitude, y = Latitude, fill = cut_number(genomic_offset_GF_LC, n = 5), size = elevation), shape = 21, color = "black") +
  scale_fill_manual(
    values = colors,
    labels = c("low values", "", "", "", "high values"),
    drop = FALSE, na.translate = FALSE
  ) +  # Ensure all levels are shown in the legend
  geom_sf(data = admin, fill = NA, size = 0.1) +
  scale_size_continuous(range = c(1, 5), name = "Elevation") +  # Adjust size range as needed
  coord_sf(xlim = c(-10, 30), ylim = c(35, 62), expand = FALSE) +
  xlab("Longitude") + ylab("Latitude") +
  guides(fill = guide_legend(title = "Genomic offset"), size = guide_legend(title = "Elevation")) +
  ggtitle("Genomic offset and Elevation across populations LC future") +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "right",
    panel.grid = element_blank(),
    strip.text = element_text(size = 11),
    plot.title = element_text(hjust = 0.5, color = "Black", face = "italic")
  )

# Print the plot
print(plot)
```
```{r include=FALSE}
#save
pdf("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/GF/figures/Genomic_offset_across_population_Taxus_elevation.pdf");print(plot);dev.off()
```
We can see that some populations in altitude have indeed high values of genomic offset but it's not true for every populations (see Greece).

More generally, we can see the same results as for RDA for the correlation between models and periods. 
For the MC LC, we have almost the same GO values regardless of the period. 
For the comparison neutral vs outliers set, we have also similar results with an overall correlation of 0.77 and 0.76. 
We can ask ourself that maybe the values are similar but the rank of population is not, we will investigate that: 


```{r Rank of outliers populations}
list_period <- c("present","future")
list_set <- c("random","LC","MC")

for(i in 1:length(list_period)){
  
  period <- list_period[i]
  
  for(x in 1:(length(list_set)-1)){
    
    set <- list_set[x]
    set2 <- list_set[x+1]
    
    #merge two dataset
GO_RDA_merge <- merge(GO_df <- get(paste0("genomic_offset_GF_",set,"_",period)),GO_df <- get(paste0("genomic_offset_GF_",set2,"_",period)),"Population")

#df
GO_RDA_set_df<- data.frame(Population= GO_RDA_merge$Population,GO_RDA_1=GO_RDA_merge[,2],GO_RDA_2=GO_RDA_merge[,3])

#RDA_GO rank
GO_RDA_set_df$rank_1 <- rank(GO_RDA_set_df$GO_RDA_1)

#GF_GO rank
GO_RDA_set_df$rank_2 <- rank(GO_RDA_set_df$GO_RDA_2)


#plot
GO_RDA_set_df_meta <- merge(GO_RDA_set_df,meta_data_pop_order[,c(1,2)],"Population")

GO_RDA_set_df_meta$Country <- as.factor(GO_RDA_set_df_meta$Country)

Scatterplot <- ggplot(GO_RDA_set_df_meta,aes(x=rank_1,y=rank_2)) +
  geom_point(aes(color=Country),size=3)+
  scale_colour_manual(name="Countries",
                      values = c("orangered3","gold2","darkorchid3","navyblue","turquoise2","green3","blue","red","black","gray","orange","darkgreen")) +
  geom_abline(intercept = 0, slope = 1, color="gray60")+
    ggtitle(paste0("Comparison GO rank of populations RDA ",set,"/",set2," ",period))+ 
    theme_set(theme_bw())

plot(Scatterplot)


pdf(paste0("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/GF/figures/comparison/Scatterplot_comparison_rank_pop_",set,"_",set2,"_",period,"_Taxus.pdf"));print(Scatterplot);dev.off()
  }
}
```

We can also look at the trend of maladaptation to see if population currently with higher predicted maladaptation will be the one in the future. 

```{r trend GO}

list_set <- c("random","LC","MC")

for(i in 1:length(list_set)){
  
  set <- list_set[i]
  
  period_1 <- c("present")
  period_2 <- c("future")
  
  df_pres <- get(paste0("genomic_offset_GF_",set,"_",period_1))
  df_fut <- get(paste0("genomic_offset_GF_",set,"_",period_2))
  
  #df
df_f_pres <- data.frame(Population = df_pres[,1], Values = df_pres[,2])
df_f_fut <- data.frame(Population = df_fut[,1], Values = df_fut[,2])

#add period
df_f_pres$Period <- "present"
df_f_fut$Period <- "Future"


#combine them
df_GO <- rbind(df_f_fut,df_f_pres)
df_GO$Period <- factor(df_GO$Period, levels = c("Present", "Future"))

#plot

trend_plot <- ggplot(df_GO, aes(x = Period, y = Values, group = Population)) +
  geom_line(aes(color = Population)) +   # Line plot, connecting the points for each population
  geom_point(aes(color = Population)) +  # Adding points for each value
  theme_minimal() +
  labs(title = paste0("Genomic offset GF trend across present and future ",set),
       x = "Period",
       y = "Genomic offset prediction") +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 11),
    plot.title = element_text(hjust = 0.5, color = "Black", face = "italic")
    )

    plot(trend_plot)
    
#save
    #pdf(paste0("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/RDA/figures/comparison/GO_pres_future/Scatterplot_comparison_trend_GO_present_future_",set,"_Taxus.pdf"));print(trend_plot);dev.off()
  
}
```

      5. Comparison GO results across methods
      
We ran GO calculation for 2 various methods, RDA and GF, using different approaches (linear vs machine learning) to estimate the GO values. We are interested in comparing if they gave similar results:
We can compare Present and Future genomic offset predictions: 

```{r correlation values GO across methods present}
list_period <- c("present","future")
list_set <- c("random","LC","MC")


for(i in 1:length(list_period)){
  
  period <- list_period[i]
  
  for(x in 1: length(list_set)){
    
    set <- list_set[x]
    
    load(paste0("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/RDA/data/Genomic_offset_Taxus_Adapcon_Gentree_RDA_",set,"_",period,".Rdata"))
    
  }
}


#df present
df_tot_GO_predictions_present <- data.frame(Population=genomic_offset_GF_LC_future[,1],GO_rand_GF=genomic_offset_GF_random_present[,2],GO_rand_RDA=Genomic_offset_Taxus_Adapcon_Gentree_RDA_random_present[,2],GO_LC_GF=genomic_offset_GF_LC_present[,2],GO_LC_RDA=Genomic_offset_Taxus_Adapcon_Gentree_RDA_LC_present[,2],GO_MC_GF=genomic_offset_GF_MC_present[,2],GO_MC_RDA=Genomic_offset_Taxus_Adapcon_Gentree_RDA_MC_present[,2])
#cor
correlation <- cor(df_tot_GO_predictions_present[,-1])

corrplot(correlation, method = "number", addrect = 2, col = c("darkorange","darkred"), type = "lower", tl.col = "black", tl.cex = 0.6, number.cex = 0.6,title="Comparison present genomic offset predictions across RDA and GF")


#df future
df_tot_GO_predictions_future <- data.frame(Population=genomic_offset_GF_LC_future[,1],GO_rand_GF=genomic_offset_GF_random_future[,2],GO_rand_RDA=Genomic_offset_Taxus_Adapcon_Gentree_RDA_random_future[,2],GO_LC_GF=genomic_offset_GF_LC_future[,2],GO_LC_RDA=Genomic_offset_Taxus_Adapcon_Gentree_RDA_LC_future[,2],GO_MC_GF=genomic_offset_GF_MC_future[,2],GO_MC_RDA=Genomic_offset_Taxus_Adapcon_Gentree_RDA_MC_future[,2])

#cor
correlation <- cor(df_tot_GO_predictions_future[,-1])

corrplot(correlation, method = "number", addrect = 2, col = c("darkorange","darkred"), type = "lower", tl.col = "black", tl.cex = 0.6, number.cex = 0.6,title="Comparison present genomic offset predictions across RDA and GF")
```


We can see that the correlation between the GO values across methods is variable. The correlation between GF and RDA for the LC set is 0.44 indicating that methods are suggesting a similar pattern but with a lot a discrepancies.
Other than that, GF LC and random have a correlation of 0.76. 
Others have no correlation of negative correlation

We can also see if the rank of population is similar between methods. Populations with a rank near 1 are the ones with the lowest GO values. 

```{r comparison rank methods}
list_period <- c("present","future")
list_set <- c("random","LC","MC")

for(i in 1:length(list_period)){
  
  period <- list_period[i]
  
  for(x in 1:length(list_set)){
    
    set <- list_set[x]

    
    #merge two dataset
GO_RDA_merge <- merge(get(paste0("genomic_offset_GF_",set,"_",period)),get(paste0("Genomic_offset_Taxus_Adapcon_Gentree_RDA_",set,"_",period)),"Population")

#df
GO_RDA_set_df<- data.frame(Population= GO_RDA_merge$Population,GO_RDA_1=GO_RDA_merge[,2],GO_RDA_2=GO_RDA_merge[,3])

#RDA_GO rank
GO_RDA_set_df$rank_2 <- rank(GO_RDA_set_df$GO_RDA_1)

#GF_GO rank
GO_RDA_set_df$rank_1 <- rank(GO_RDA_set_df$GO_RDA_2)


#plot
GO_RDA_set_df_meta <- merge(GO_RDA_set_df,meta_data_pop_order[,c(1,2)],"Population")

GO_RDA_set_df_meta$Country <- as.factor(GO_RDA_set_df_meta$Country)

Scatterplot <- ggplot(GO_RDA_set_df_meta,aes(x=rank_1,y=rank_2)) +
  geom_point(aes(color=Country),size=3)+
  scale_colour_manual(name="Countries",
                      values = c("orangered3","gold2","darkorchid3","navyblue","turquoise2","green3","blue","red","black","gray","orange","darkgreen")) +
  geom_abline(intercept = 0, slope = 1, color="gray60")+
  labs(title = paste0("Comparison GO rank of populations GF RDA ",set," ",period),
       x = paste0("RDA ",set),
       y = paste0("GF ",set)) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 11),
    plot.title = element_text(hjust = 0.5, color = "Black", face = "italic")
    )

plot(Scatterplot)


pdf(paste0("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/Genomic_offset/comparison/scatterplot/Scatterplot_comparison_rank_pop_RDA_GF_",set,period,"_Taxus.pdf"));print(Scatterplot);dev.off()
  }
}
```


