---
title: "New_processing_climatic_data_april_T_Adapcon_Gentree"
author: "Thomas Francisco"
date: "2024-04-29"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)

rm(list = ls())

library(dplyr) 
library(tidyverse)
library(ggplot2)
library(corrplot)  
library(dismo)
library(raster)
library(vegan)
library(plyr)
library(writexl)
```

```{r meta data}
meta_data_pop <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Populations/taxus_sample_29pop.csv",h=T,sep=";",dec=",")
meta_data_vcf <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/samples/samples_taxus_baccata_adapcon_gentree.csv",h=T,sep=";",dec=",")
```

This script is a pipeline to process the climatic/environmental data from the row output of *ClimateDT* (or any program that would give data raster tif) to a dataframe with center and reduce the choosen variables for past/present and future climatic data.
This script is separated in multiples steps:
- Pre-selection of climatic variables and visualization between pop (see if adaptive gradient could emerge from )
- Loading climatic data and processing them
- Selection of climatic variables: explained genetic variance, check multicollinearity between variables and VIF
- Comparison of the values of the different climatic models for the future between populations 
- Standardization of the retained climatic values


**1. Pre-selection of climatic variables**
  The first step consist of going through the literature of the targeted species and define a set of climatic variables potentially drivers (inducing selective pressure) of adaptive genetic variation across the species range.  
The goal is to retain biological informative climatic variables only. This enable to reduce the set of tested climatic variables and also enable to work on the hypothesis before starting the analysis.
The goal is not to retain only 2 or 3 variables but also not to retain 40 of them. This pre-selection could retain 10-15 variables depending on the species and the richness of the literature related to it.

For Taxus, we retained as a first pre selection: **Bio1,2,4,7,12,15,AHM, precip June_July, precip Feb_to_July, Temperature June to September, Min temperature Novembre to January** and now we're applying a less conservative pre-selection of variables: **Bio 1,2,3,4,5,6,7,9,10,12,13,14,15,16,17,18,MGSP,AHM,SHM**.

  
  **2. Loading climatic data and processing them**
  Climatic data comes from **ClimateDT** in two different types: layer of points and raster.
We extracted the data from the past and from the future. One major differences between past and future data results in the shape of the data: 
  - Past data are available in 1 value per year for each variables (for each populations/pixel selected). As an example, for the period 1901-1950, we have 50 values for each variables for each populations
- Future data are not so accurate, we only have the mean of the values for the selected period. As a example, for the period 2041-2070, we have 1 value for each variables for each populations. -> Not sure about this because I have 1 value per year per pop even for the future data...

One of the issues is that all the bioclimatic variables (bio) or other related climatic variables are build using monthly values of tmin, tmax and precipitation. We cannot just perform the mean of the 50 values of bio 1 from 1901 to 1950 to obtain the mean value of bio 1 for this period. We need to average the tmin, tmax and precipitation for each month to have 1 value of tmin for January for the period 1901-1950, 1 value of tmin for February for the period 1901-1950 ... 
Then, we can use the **biovars** function from the **Dismo** R package to calculate the 19 bioclim variables for the period of interest (e.g 1901-1950).
To calculate the other climatic variables such as AHM for the same specific period, we can follow the procedure of ClimateDT by using the new Bio1,Bio10 and Bio12... obtained from biovars.
For other variables not directly linked to tmin, tmax, precipitation or bioclim variables such as GDD5, the average value for a specific period is not straightforward to get, data are coming from Maurizio Marchi (IBBR-CNR).

      -> Here we already check for the period in the previous script and test for differences between layers of points and raster.
      
So based on that, **we will use the 1901-1950 period and we will use raster data because the differences are important and for future data  we need to use the rasters because we want different predictions from several models (and Maurizio never explained why despite asking for it 3-5 times).** 

      - Load the past climatic data

```{r load raster data}
#create an R object where all the climatic rasters are.
ras.bio <- stack(list.files("C:/Users/tfrancisco/Documents/Thèse/Data/Climatic/1901-1950_raster/", pattern = ".tif$", full.names = T)) 
#names of the raster (because they are not in the classic order Bio 1, 2 etc they are order by number so bio 1, 10, 11 etc)
names(ras.bio) <- unlist(strsplit(unlist(lapply(strsplit(list.files("C:/Users/tfrancisco/Documents/Thèse/Data/Climatic/1901-1950_raster/", pattern = ".tif$", full.names = T), split = "C:/Users/tfrancisco/Documents/Thèse/Data/Climatic/1901-1950_raster/"), function(x) x[2])), split = ".tif"))


#we extracted the climatic values for each populations from the raster based on their coordinates
coords<- data.frame(apply(meta_data_pop[,c(5:4)], 2, as.numeric))#we need to have longitude then latitude not the other way around.
past_climatic_data_raster <- data.frame(meta_data_pop$Population,raster::extract(ras.bio, coords))#attention si le package tidyr est actif, cela masque la fonction extract du package raster
colnames(past_climatic_data_raster)=c("Population",names(ras.bio))

#add country info 
past_climatic_data_f <- merge(meta_data_pop[,c(1,2)],past_climatic_data_raster, "Population")

```

      - Variation of the variables between populations
      
```{r violin plot in a loop, eval=FALSE, include=FALSE}
#do the violin plot in a loop for each climatic variables

#input data for the loop
data <- past_climatic_data_f
regions <-past_climatic_data_f$Country
bioclim_vars <- c("bio1", "bio2", "bio3", "bio4","bio5","bio6","bio7","bio9", "bio10","bio12","bio13","bio14","bio15","bio16", "bio17", "bio18","AHM", "MGSP", "SHM", "GDD5")
unit_var <- c("°c","°c","index","°c","°c","°c","°c","°c","°c","mm","mm","mm","index","mm","mm","mm","°c/mm","mm","°c/mm","°c*days" )

#results: the loop
for (x  in 1:length(bioclim_vars)) {
  var <- bioclim_vars[x]
   unit <- unit_var[x]
  data_subset <- data.frame(data[, paste0(var)]) #for each variable in bioclim_vars, we extract the two periods and created a data.frame with only these two variables. 
  colnames(data_subset) <- c("var1")
  # Create scatterplot to compare the bioclimatic variable between the two periods
  plot <- ggplot(data_subset, aes(x =var,y= var1)) +#var is the values and var1 is just the name of the variable 
  geom_violin(trim = FALSE) +
  geom_point(aes(color = regions), shape = 16, position = position_jitter(seed = 1, width = 0.2)) +
  labs(colour="Regions",x="variable",y=paste0("values ","(",unit,")"),title=paste0("Violin Plot of ",var," differences across populations"))+
  theme(plot.title = element_text(hjust = 0.5))
  # Print the plot
  print(plot)
  # Add any additional customization or saving the plot if needed
png(paste0("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/climate/past_climatic_data/violin_plot/violin_plot_",var,".png"));print(plot)
;dev.off()
}
 
  ######"
    #If we want to add values from different climatic models, we need to do a thing like thats: data_violin <- gather(revelant_variables_graph,key="revelant_variables",value="Value",-Population,-Country)#create a dataframe for the violin graph with all the values of climatic variables in 1 columns: Value and all the different coordinates for populations for the different variables in row.
  #the different values of bio1 for example from different models should be in the same dataframe -> 1 data frame per bioclimatic variables -> maybe we can do a loop for that too. 

```

Interpretation:...
Conclusion: ...



  
  **3. Selection of climatic variables**
  
  Selection of climatic variables will be in 2/3 steps: 
  
    i. Difference between past and future values

We will not do this step because if the variables are not changing into the future that is also an information that the climate change will not impact much this species or these populations if some of them are very locally adapted to a variable that will not change. 



      ii. Imprecision along studied area
Accuracy of the climatic variables is not homogeneous along the studied area and also along the climatic variables.
If layers of uncertainty are available, we could check if some variables have a high inaccuracy along the studied area and then remove them. 
```{r Imprecision along the studied area}

```
Interpretation:
Conclusion:

 -> NOT POSSIBLE, MAURIZIO Never answer about that
 
      iii. Explained variance
In this step, we want to see which variables explained the most the genetic variation. 
To do this, we will perform a RDA
We need the genomic data
```{r genomic data}
#genomic data
load("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/genetic_new/data_allelic_frequencies_29pop_adapcon_gentree_475_8616.Rdata")

genomic_matrix <- data_allelic_frequencies_29pop_adapcon_gentree_475_8616
```
We also need to scale the climatic var: 
```{r scale}
past_climatic_data_scale <- past_climatic_data_f[,-c(1,2)]%>%
  apply(2,as.numeric) %>% as.data.frame %>% 
mutate(across(where(is.numeric), scale))
```

We can perform the RDA
```{r RDA}
#null model
RDA_null <- rda(genomic_matrix~1,past_climatic_data_scale )

#model with all the variables
RDA_full <- rda(
formula=genomic_matrix~bio1+bio2+bio3+bio4+bio5+bio6+bio7+bio9+bio10+bio12+bio13+bio14+bio15+bio16+bio17+bio18+AHM+MGSP+SHM+GDD5,data = past_climatic_data_scale,scale=F)

#stepwise by selecting only variables that increase the adjusted R^2 of the model and that have a pvalue lower than 0.05
Variable_selection_pv0.05 <- ordiR2step(RDA_null, RDA_full, Pin = 0.05, R2permutations = 1000, R2scope = T)

Variable_selection_pv0.05$anova

#we want, following Archambeau et al 2024, investigate if the permutation stepwise identified the same variables of interest depending on the run:
#we run 20 independent stepwise models.
nbmodels <- 20

# Stepwise selection with ordiR2step function
rep20_variable_selec_pv0.05 <- lapply(1:nbmodels, function(x) {
  mod <- ordiR2step(RDA_null, RDA_full, Pin = 0.05, R2permutations = 1000, R2scope = T)
  return(names(mod$CCA$envcentre))}) %>% 
  setNames(paste0("model",1:nbmodels)) %>% 
  ldply(function(x) data.frame(variables=x),.id="models") %>%  
  dplyr::summarise(count(variables)) %>% 
  setNames(c("variable","count"))


rep20_variable_selec_pv0.05

```
Interpretation: Bio 2, 4, 7 and 9 seems to be important. Bio 4 and 9 are identified as relevant in 20 of the 20 runs and Bio 2 and 7 in 19 of the 20. 
Conclusion: According to the stepwise analysis, BIO 2, 4, 7 and 9 seems to be of interest to explain the genetic variation. 
      

      iv. Multicollinerity between variables
Here, we want to select the variables with low collinearity between them.
To do so, we decided to keep only the variables not too correlated (<absolute corr of 0.75) from the remaining variables after step 1.

```{r multicollinearity between variables, eval=FALSE, include=FALSE}
#function to do it
#matrix of correlation
correlation_function <-function(data,threshold){
    data_correlation <- subset(data,select= -c(Population,Country)) 
      rownames(data_correlation) <- data$Populations
      correlation <- cor(data_correlation)
  correlation[abs(correlation) <= threshold] <- 0
corr_plot <- corrplot(correlation, method = "number", addrect = 2, col = c("red", "white", "red"), type = "lower", tl.col = "black", tl.cex = 0.6, number.cex = 0.6)
}

#right order
past_climatic_data_order <- past_climatic_data_f[,c(1,2,4,13:19,5:12,3,20,21,22)]

#correlation past/present
data_present <- past_climatic_data_order #we use the dataset without the variables we discarded previously
threshold <- 0.75
correlation_past <-correlation_function(data_present,threshold)



```
Conclusion: Based on the correlation between variable and step 3, we can select **BIO 1, 2, 3, 4, 5, 9, 12, 15**

      v. VIF: 
      
We want to test the inflation factor between variables retained to be sure that we do not take overcorrelated variables

```{r VIF}
#subset step 4
RDA_subset_step4 <- rda(formula=genomic_matrix~bio1+bio2+bio3+bio4+bio5+bio9+bio12+bio15,data = past_climatic_data_scale,scale=F)
anova.cca(RDA_subset_step4)
RsquareAdj(RDA_subset_step4)
##vif
vif.cca(RDA_subset_step4)

#adjsuted subset based on VIF
RDA_adjusted_subset <- rda(formula=genomic_matrix~bio1+bio2+bio4+bio9+bio12+bio15,data = past_climatic_data_scale,scale=F)
anova.cca(RDA_adjusted_subset)
RsquareAdj(RDA_adjusted_subset)
##vif
vif.cca(RDA_adjusted_subset)

```
Interpretation: We can see that the VIF of some variable in the subset coming from step 4 is too high (over 160 for bio2). Based on that and the information that some variables are of interest to explain the genetic variation, we adjusted the subset to lower the VIF (below 10, all the final retained variables have a VIF < 3)
Conclusion: we retained the subset: **BIO 1, 2, 4, 9, 12, 15**
   
  **5. Comparison of the future retained climatic values between populations**
  
  Multiple climatic models that predict the future climate are available. Choosing one of them could be a challenge because sometimes the difference in prediction are important. To avoid bias by selecting only one climatic model that could be an extreme, we selected 5 of them for the future data. 
We want to compare them to see if there is a lot a differences in predictions between models. 
To do so, we performed a ... see paper of benjamin or felix showing that
```{r Comparison of the climatic models prediction}
#look above,I explained how we can modify the loop of violin plot to compare the values of predictions for the different climatic models into the future 
```

If, we want to add the values for bio1 and other bio from different climatic models, we need to do like in the function violin plot by creating a columns with the nameof the different models and the values in the same columns. 

```{r plot of the variability of variables across pop}

violin_plot <- function(revelant_variables_graph,ID,regions){#may be interesting to do it by regions and not populations to better vizualise
data_violin <- gather(revelant_variables_graph,key="revelant_variables",value="Value",-Population,-Country)#create a dataframe for the violin graph with all the values of climatic variables in 1 columns: Value and all the different coordinates for populations for the different variables in row.
data_violin$Country=as.factor(data_violin$Country)

p <- ggplot(data_violin, aes(x = revelant_variables, y = Value)) +
  geom_violin(trim = FALSE) +
  geom_point(aes(color = Country), shape = 16, position = position_jitter(seed = 1, width = 0.2)) +
  labs(colour="Regions",x="climatic variables",y="Difference between future and past climate",title="Violin Plot of Climatic Factors by Regions")+
  theme(plot.title = element_text(hjust = 0.5))
return(p)
}


revelant_variables_graph <- relative_data_difference#selection of ID, populations and the climatic variables
Groups <- past_climatic_data_1901_1950$Population
regions <-past_climatic_data_1901_1950$Country
  
s=violin_plot(revelant_variables_graph,ID,regions)
```

**6. Standardization of the retained climatic values**
  
 The final step is to create a data frame with the retained climatic variables standardized for further analysis. 
We standardized the past and future data with the same values to enable comparison of index between past and future based on the climatic data (e.g Adaptive index, genomic offset...)

```{r normalize the data, message=FALSE, warning=FALSE}
#creation of the scaled matrix

Past_new_6_Climatic_data_scale <-past_climatic_data_order[,c(3,4,6,10,12,15)] %>% 
    scale()


Past_new_6_Climatic_data_scale_df <- data.frame(past_climatic_data_order[,c(1,2)],Past_new_6_Climatic_data_scale)

colnames(Past_new_6_Climatic_data_scale_df) <- c("Population","Country","Annual_Tc","Diurnal_range_Tc","Tc_Seasonality","Tc_driest_quarter","Annual_P","P_Seasonality")


  scale_env_value_new_cli <- attr(scale(Past_new_6_Climatic_data_scale_df[,-c(1,2)]), 'scaled:scale')
center_env_value_new_cli <- attr(scale(Past_new_6_Climatic_data_scale_df[,-c(1,2)]), 'scaled:center') 

#Save the matrix and the normalized values

write_xlsx(Past_new_6_Climatic_data_scale_df,"C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Climatic_data/new_selection/Past_new_6_Climatic_data_scale_df.xlsx")
save(Past_new_6_Climatic_data_scale_df,file="C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Climatic_data/new_selection/Past_new_6_Climatic_data_scale_df.Rdata")
save(scale_env_value_new_cli,file="C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Climatic_data/new_selection/scale_env_value_new_cli.Rdata",force=T)
save(center_env_value_new_cli,file="C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Climatic_data/new_selection/center_env_value_new_cli.Rdata",force=T)

```

We also for LFMM analysis, have to do a dataframe with climatic values at the individual level: with individuals of the same pop having the same climatic values -> redundancy. 

```{r new climatic matrix at individual-level}
#load meta_data at individual level

climatic_data_indivdual_level_new_selection <- merge(meta_data_vcf[,-2],past_climatic_data_order[,c(1:4,6,10,12,15)],"Population")

climatic_data_indivdual_level_scaled_new_selection <- climatic_data_indivdual_level_new_selection[,-c(1:3)] %>% 
  scale()

climatic_data_indivdual_level_scaled_new_selection_df <- data.frame(climatic_data_indivdual_level_new_selection[,c(1:3)],climatic_data_indivdual_level_scaled_new_selection)

#rename columns
colnames(climatic_data_indivdual_level_scaled_new_selection_df) <- c("Population","VCF_ID","Country","Annual_Tc","Diurnal_range_Tc","Tc_Seasonality","Tc_driest_quarter","Annual_P","P_Seasonality")

write_xlsx(climatic_data_indivdual_level_scaled_new_selection_df,"C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/climatic_data_indivdual_level_scaled_new_selection_df.xlsx")
save(climatic_data_indivdual_level_scaled_new_selection_df,file="C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Climatic_data/new_selection/climatic_data_indivdual_level_scaled_new_selection_df.Rdata")

```