---
title: "Variance_partitioning_new_variables"
author: "Thomas Francisco"
date: "2024-04-30"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(cache = FALSE)

library(vegan)
library(dplyr)
```

The goal of this script is to perform a variance partitioning to disentangle the variables explaining the genetic variation. The second goal is to see if the climatic variables explained some of the genomic variation suggesting that we could find some loci candidates of local adaptation to climatic variables across populations. To perform this variance partitioning, we used the RDA and pRDA analysis. These analysis are explained in the scripts RDA_candidate_detection and pRDA_candidate_detection. 

First, we load the datasets. Because the goal is to disentangle the effects of the ancestry, climate and IDB on the genetic variation, we used proxies for ancestry and IBD.

```{r intial data}

#climatic data
##0.6 corr
Past_climatic_new_var <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Climatic_data/new_selection/Past_new_6_Climatic_data_scale_df.csv",sep=";",dec=",")

#Ancestry proxies
##PCs axis
structure_PCs <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/GEA/Structure_proxy/PCa_score_T_adapcon_gentree_scale_new.csv",sep=";",dec=",")

##STRUCTURE Groups
load("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/GEA/Structure_proxy/Cluster_score_STRUCTURE_T_Adapcon_gentree.Rdata")
Structure_cluster <- Cluster_score_STRUCTURE_T_Adapcon_gentree

#IBD proxy
#dbMEMs
IBD_dbMEMs <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/GEA/dbMEMs/meta_data_dbmem_29pop_adapcon_gentree_scale.csv",sep=";",dec=",")


#genomic data
load("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/genetic_new/data_allelic_frequencies_29pop_adapcon_gentree_475_8616.Rdata")

genomic_matrix <- data_allelic_frequencies_29pop_adapcon_gentree_475_8616
```


```{r merge all informations into 1 table}
#new_var
Data_RDA <- merge(Past_climatic_new_var,structure_PCs[,c(1:3)],"Population") %>% merge(Structure_cluster,"Population") %>% merge(IBD_dbMEMs[,c(2,4,5,7:10)],"Population") 

#scale coord
scale_coord <- Data_RDA[,c(14,15)] %>%
  apply(2,as.numeric) %>% scale()

Climatic_data_RDA_pRDA <- data.frame(Data_RDA[,-c(14,15)],scale_coord)

#Save for RDA, pRDA
save(Climatic_data_RDA_pRDA,file="C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/GEA_new_var/variance_partitioning/Climatic_data_RDA_pRDA.Rdata")
```

One assumption of the GEA models is that the environmental variables explains some of the genetic variation. We want to test that.   
Also, the variance partitioning enables to visualize if there is a lot a covariation between the structure, IBD (geo) and environment.  
Because there is not one good proxy of ancestry and IBD, we used several of them for population structure and geography:

    - coords as IBD proxy and PCs as ancestry proxy

```{r variance partitionning with coord as IBD proxy and PCs as ancestry proxy}
#total model
M_tot_coord <- rda(formula=genomic_matrix~Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality+PC1+PC2+Longitude+Latitude,data = Climatic_data_RDA_pRDA,scale=T)
#RsquareAdj(M_tot_coord)
#anova.cca(M_tot_coord)


##partial_ancestry_model
M_ancestry_coord <- rda(formula = genomic_matrix ~ PC1+PC2 +Condition(Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality +Longitude+Latitude), data = Climatic_data_RDA_pRDA,scale=T)
#RsquareAdj(M_ancestry_coord)
#anova.cca(M_ancestry_coord)


#partial_geo_model
M_geo_coord <- rda(formula = genomic_matrix ~ Longitude+Latitude+ Condition(Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality+PC1+PC2), data = Climatic_data_RDA_pRDA,scale=T)
#RsquareAdj(M_geo_coord)
#anova.cca(M_geo_coord)

#partial_env_RDA
M_IBE_coord <- rda(formula = genomic_matrix ~  Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality +Condition(PC1+PC2+Longitude+Latitude) , data = Climatic_data_RDA_pRDA,scale=T)
#RsquareAdj(M_IBE_coord)
#summary(M_IBE_coord)
#anova.cca(M_IBE_coord)

```
Interpretation:   
R^2 tot     : 0.52                     Adj R^2 tot     : 0.25    
R^2 ancestry: 0.13                     Adj R^2 ancestry: 0.10  
R^2 IBD     : 0.06                     Adj R^2 IDB     : 0.01  
R^2 climate : 0.18                     Adj R^2 climate : 0.02  

    - dbMEMs as IBD proxy and Structure groups as ancestry proxy

```{r variance partitionning with mscore_structure as Structure proxy and dbMEMs as IBD proxy}
#total model
M_tot_db <- rda(formula=genomic_matrix~Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality+Group1+Group2+Group3+MEM1+MEM2+MEM3+MEM4,data = Climatic_data_RDA_pRDA,scale=T)
#RsquareAdj(M_tot_db)
#anova.cca(M_tot_db)


##partial_ancestry_model
M_ancestry_db <- rda(formula = genomic_matrix ~ Group1+Group2+Group3 +Condition(Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality +MEM1+MEM2+MEM3+MEM4), data = Climatic_data_RDA_pRDA,scale=T)
#RsquareAdj(M_ancestry_db)
#anova.cca(M_ancestry_db)


#partial_geo_model
M_geo_db <- rda(formula = genomic_matrix ~ MEM1+MEM2+MEM3+MEM4+ Condition(Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality+Group1+Group2+Group3), data = Climatic_data_RDA_pRDA,scale=T)
#RsquareAdj(M_geo_db)
#anova.cca(M_geo_db)



#partial_env_RDA
M_IBE_db <- rda(formula = genomic_matrix ~  Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality +Condition(Group1+Group2+Group3+ MEM1+MEM2+MEM3+MEM4) , data = Climatic_data_RDA_pRDA,scale=T)
#RsquareAdj(M_IBE_db)
#summary(M_IBE_db)
#anova.cca(M_IBE_db)

```

Interpretation:   
R^2 tot     : 0.59                     Adj R^2 tot     : 0.24    
R^2 ancestry: 0.11                     Adj R^2 ancestry: 0.04  
R^2 IBD     : 0.17                     Adj R^2 IDB     : 0.08  
R^2 climate : 0.22                     Adj R^2 climate : 0.08  

    - coords as IBD proxy and Structure groups as ancestry proxy

```{r variance partitionning with coord as IBD proxy and Structure as ancestry proxy}
#total model
M_tot_coord <- rda(formula=genomic_matrix~Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality+Group1+Group2+Group3+Longitude+Latitude,data = Climatic_data_RDA_pRDA,scale=T)
RsquareAdj(M_tot_coord)
#anova.cca(M_tot_coord)


##partial_ancestry_model
M_ancestry_coord <- rda(formula = genomic_matrix ~ Group1+Group2+Group3 +Condition(Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality +Longitude+Latitude), data = Climatic_data_RDA_pRDA,scale=T)
RsquareAdj(M_ancestry_coord)
#anova.cca(M_ancestry_coord)


#partial_geo_model
M_geo_coord <- rda(formula = genomic_matrix ~ Longitude+Latitude+ Condition(Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality+Group1+Group2+Group3), data = Climatic_data_RDA_pRDA,scale=T)
RsquareAdj(M_geo_coord)
#anova.cca(M_geo_coord)



#partial_env_RDA
M_IBE_coord <- rda(formula = genomic_matrix ~  Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality +Condition(Group1+Group2+Group3+Longitude+Latitude) , data = Climatic_data_RDA_pRDA,scale=T)
RsquareAdj(M_IBE_coord)
#summary(M_IBE_coord)
#anova.cca(M_IBE_coord)

```
Interpretation:   
R^2 tot     : 0.51                     Adj R^2 tot     : 0.20    
R^2 ancestry: 0.12                     Adj R^2 ancestry: 0.12  
R^2 IBD     : 0.09                     Adj R^2 IDB     : 0.05  
R^2 climate : 0.21                     Adj R^2 climate : 0.05  


    - dbMEMs as IBD proxy and PCs as ancestry proxy
    
```{r variance partitionning with dbMEMs as IBD proxy and PCs as ancestry proxy}
#total model
M_tot_db <- rda(formula=genomic_matrix~Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality+PC1+PC2+MEM1+MEM2+MEM3+MEM4,data = Climatic_data_RDA_pRDA,scale=T)
RsquareAdj(M_tot_db)
#anova.cca(M_tot_db)


##partial_ancestry_model
M_ancestry_db <- rda(formula = genomic_matrix ~ PC1+PC2 +Condition(Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality +MEM1+MEM2+MEM3+MEM4), data = Climatic_data_RDA_pRDA,scale=T)
RsquareAdj(M_ancestry_db)
#anova.cca(M_ancestry_db)


#partial_geo_model
M_geo_db <- rda(formula = genomic_matrix ~ MEM1+MEM2+MEM3+MEM4+ Condition(Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality+PC1+PC2), data = Climatic_data_RDA_pRDA,scale=T)
RsquareAdj(M_geo_db)
#anova.cca(M_geo_db)



#partial_env_RDA
M_IBE_db <- rda(formula = genomic_matrix ~  Annual_Tc+Diurnal_range_Tc+Tc_Seasonality+Tc_driest_quarter+Annual_P+P_Seasonality +Condition(PC1+PC2+ MEM1+MEM2+MEM3+MEM4) , data = Climatic_data_RDA_pRDA,scale=T)
RsquareAdj(M_IBE_db)
#summary(M_IBE_db)
#anova.cca(M_IBE_db)

```
Interpretation:   
R^2 tot     : 0.59                     Adj R^2 tot     : 0.27    
R^2 ancestry: 0.10                     Adj R^2 ancestry: 0.08  
R^2 IBD     : 0.13                     Adj R^2 IDB     : 0.04  
R^2 climate : 0.18                     Adj R^2 climate : 0.03  


**Conclusion: we will retain the model with dbMEMs as IBD and PCs as ancestry proxy because that is one the model that explain the most the genetic variation while having a low number of variables.** 


